Thermal reactionomes reveal divergent responses to thermal extremes in warm

### 2016-01-12 doc/ApTranscriptome_manuscript.Rmd
edits for BMC Genomics

**Results** We found that at least 2% of all genes were thermally responsive. The majority of upregulation was specific to exposure to low temperatures. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than did *A. carolinensis*, consistent with the *enhanced response* hypothesis. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea*, and required more extreme temperatures to induce down-regulation in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.

@@

We use the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* (part of the *A. rudis*-species complex, B. DeMarco personal communication) and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences a higher mean annual temperature (MAT) (14.6°C) and less seasonal temperature variation (temperature seasonality = 7,678°) than does *A. picea* (MAT = 4.6°C, seasonality = 10,008°; [@hijmans2005]) at their respective collection sites. In controlled laboratory experiments, these warm- and cold-climate species exhibit corresponding differences in their critical maximum (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*). These differences between species in their thermal environments and physiological tolerances allowed us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural changes [@lockwood2012] or behavioral adaptations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways [@neelakanta2012; @franssen2011].

Finally, the *genetic assimilation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify if, and under what conditions, these mechanisms contribute to thermal adaptation.

@@

The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts. About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. We fit polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship. To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationshp by chance alone. We retained the `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) true positive transcripts that exceeded null expectation from the resampling analysis. Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

@@

In addition, we also examined the specific patterns of shifts from one expression category to another between species. When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 319, P < 0.001, Fig. S1). Specifically, the two species differed significantly in expression pattern, which captures differences in slope as well as category, for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts. Most of these shifted from one to another expression category.

The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig. 1). The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig. 1, left). In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig. 1, right). We tested these predictions by examining if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A. carolinensis*, and warm temperatures for the cool-climate *A. picea*) displayed the same response profile in the species that more frequently experiences those conditions.

Transcripts upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from other expression categories in *A. picea* (Fig. 2A), suggesting shared response pathways as predicted by the *enhanced response* hypothesis. In contrast, transcripts upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) shifted expression categories in *A. carolinensis* (Fig. 2B), primarily to the **Intermediate** category (Fig. 2B). These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.

@@

The gene set enrichment analysis revealed a number of gene groups enriched in each expression category. Across both species, there were 9 terms enriched in the **Bimodal** category, including terms involved in stress response (regulation of cellular response to stress, signal transduction by p53 class mediator), cell death (apoptotic signaling pathway) and cellular organization (e.g. protein complex localization). The 6 terms enriched in the **Low** category suggest that proteins undergo structural (e.g protein acylation) and organizational (single-organism organelle organization) changes to tolerate colder temperatures, possibly to maintain membrane fluidity [@ohtsu1998]. The **High** category included only a single enriched GO term, "nicotinamide metabolic process", while the **Intermediate** category had 5 terms including DNA packaging and metabolic process terms.

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *enhanced response* but not the *tolerance hypothesis*, the temperature of induction at low temperatures was significantly higher for the cool-climate *A. picea* than for *A. carolinensis* (`r round(T_on.low.ttest$estimate[1], 1)`°C) than *A. picea* (`r round(T_on.low.ttest$estimate[2], 1)`°C; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 3A), though the temperature of induction did not differ between species for high temperatures (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`).

Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Consistent with the *tolerance hypothesis*, *A. carolinensis* had greater inertia of gene expression in response to increasing temperatures. The temperature of downregulation for **Intermediate** transcripts was `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`). The difference between species was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 3B).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C). If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A. picea* were not expressed at higher baseline levels in the warm-climate *A. carolinensis* (Fig. 4A). Similarly, transcripts more upregulated at low temperatures in *A. carolinensis* did not show higher baseline levels in *A. picea* (Fig. 4B). In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001). In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (top-row of Fig. 2B). These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.

@@

Under the *enhanced response* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, *A. picea* upregulated responsive transcripts at slightly less extreme temperatures (Fig. 3A). Moreover, the transcripts upregulated in *A. picea* included about about half (`r round(tt2["Low", "Low"] / colSums(tt2)["Low"], 2)*100`%) those upregulated in *A. carolinensis* as well as an additional set of `r round(rowSums(tt2)["Low"] - tt2["Low", "Low"], 2)` transcripts (Table 1), enriched for metabolism, organization and translation processes (Table S1). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

@@

The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts. This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013]. However, in a short-term selection experiment for heat tolerance, Sikkink et al. [-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance. Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair. Whether a particular taxon evolves one strategy over another may be related to availability of alternative mechanisms as well as the the intensity, frequency and duration of temperature stress in a given environment.

Given the differences in the patterns of thermal responsiveness between species (Fig. 2), it is worth noting a number of similarities. In both species, there were 2 -- 3 times more transcripts upregulated at low than high temperatures (Table 1). The degree of upregulation at low temperatures is surprising given previous studies [@teets2012-1; @vesala2012] that found little transcriptional activity at low temperatures. However, these studies exposed organisms to a few extreme (-10 -- 0°C) temperatures. At these extremes, we also found few upregulated transcripts (Fig. 3A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig. 3). A potential explanation for this pattern is that increased gene expression functions to support elevated metabolism at moderately cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002]. The observation that more transcripts were upregulated at low than high temperatures could also be due to stronger selection on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.

Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression in this study. Only ~10% of transcripts upregulated in response to temperature were bimodal, and for both activation and down-regulation, thresholds differed between species at only one temperature extreme (Fig. 3). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments.

A major contribution of this study is the construction of a reactionome for gene expression data (see Hodgins-Davis et al [-@hodgins-davis2012] for a similar approach), which revealed quantitative patterns of temperature response not captured in categorical comparisons. For example, the degree of upregulation at cool (~10°C) but not extreme cold temperatures was missed in previous studies that focused on extreme cold limits, as discussed above. Further, a number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" interpretations [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. Although we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.

Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. Although we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig. 2, Fig. S1). Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression -- as opposed to identifying lists of differentially-expressed transcripts -- can provide meaningful insight on the process of evolution. It should be noted, however, that although including non-linear relationships between expression and temperature captured a significantly larger range of biologically-relevant responses, it also led to a substantial increase in false positives. Empirical estimation of these rates via randomization tests, combined with robust sampling designs, can help to minimize this bias and focus results on biologically-meaningful gene sets.

@@

@@

Table S1 contains the results of the gene set enrichment analysis, showing the enriched gene ontology terms for each species in each thermal response category.

@@

The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available at the Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-39, and hf113-40 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).

@@

![Fig. 1](../results/predictions_Fig1.png)

![Fig. 2](../results/matched_observations_Fig2.png)

>>>>

**Results** We found that at least 2% of all genes changed expression with temperature. The majority of upregulation was specific to exposure to low temperatures. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than did *A. carolinensis*, consistent with the *enhanced response* hypothesis. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea*, and required more extreme temperatures to induce down-regulation in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*.

@@

In this study, we used the reactionome method to identify genes that are thermally responsive in two closely-related eastern North American ant species, *Aphaenogaster carolinensis* [@demarco2015] and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ants in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. Ants, and ectotherms in general, have little or no thermal safety margin [@sunday2014] and thus are highly susceptible to climate change [@diamond2011; @kingsolver2013], putting at risk important ecosystem services [@toro2012]. Growth chamber studies have demonstrated that reproduction of *Aphaenogaster* will be compromised by increased tempreatures [@pelini2012], while field studies [@diamond2012] and simulations [@bewick2014] indicate that ant species persistence will depend on combinations of physiology and species interactions. *Aphaenogaster carolinensis* experiences a higher mean annual temperature (MAT) (14.6°C) and less seasonal temperature variation (temperature seasonality = 7,678°) than does *A. picea* (MAT = 4.6°C, seasonality = 10,008°; [@hijmans2005]) at their respective collection sites. In controlled laboratory experiments, these warm- and cold-climate species exhibit corresponding differences in their critical maximum (44.7°C for *A. carolinensis* versus 41.3°C for *A. picea*; see Methods) and minimum temperatures (6.1°C for *A. carolinensis* versus -0.1°C for *A. picea*). These differences between species in their thermal environments and physiological tolerances allowed us to investigate adaptation to both lower and upper thermal extremes in this system.

@@

Second, the *tolerance hypothesis* [@cowles1939; @fields2001] proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from an induced response in favor of structural changes [@lockwood2012] or behavioral adaptations [@gordon2013; @sunday2014]. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways [@neelakanta2012; @franssen2011].

Finally, the *genetic assimilation hypothesis* [@waddington1953; @sikkink2014] proposes that exposure to more extreme stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, whereas transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*.

To summarise, in this project we generated the transcriptomes of two closely-related temperate ant species, and quantified their gene expression across a wide range of thermal conditions. We used a reactionome approach to identify all transcripts that changed expression across a thermal gradient each species. We then evaluated three non-mutually exclusive hypotheses (enhanced response, tolerance and genetic assimilation) of the evolution of thermal adaptation by comparing differences between the thermal reactionomes of each species. Finally, we used gene ontology information to determine which gene products and pathways are involved in thermal adaptation in the two species.

@@

The combined *Aphaenogaster* transcriptome assembly contained `r prettyNum(nrow(annotation.table), big.mark =",")` transcripts. About half of these (`r prettyNum(nrow(annotation.table.responsive), big.mark =",")`) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. We performed a BUSCO analysis [@simao2015] to assess the quality of the transcriptome assembly against the arthropod Benchmarking Universal Single-Copy Orthologs (BUSCOs). This analysis revealed that transcriptome is largely complete, as we recovered 1,426 complete single-copy BUSCOs (62%) and an additional 435 fragmented BUSCOs (16%), which is in line with results of Simao et al. [-@simao2015] for transcriptomes of other non-model species. Moreover, only 8% of the BUSCOs were found to be duplicated in the transcriptome, which indicates that while we assembled the transcriptomes of the two species together, the steps we took to collapse homologs were successful.

We quantified gene expression using the program `Sailfish` [@patro2014], and fit polynomial regression models to the expression values of each transcript to identify those that had a linear or quadratic relationship (Table S1). To account for multiple tests, we both applied a False Discovery Rate (FDR) correction, and performed a resampling analysis to determine the number of transcripts that would be expected to have a significant relationship by chance alone. We retained the `r nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)` (`r round((nrow(pre.signif.transcripts) - quantile(D$num_signif, probs = 0.95)) / nrow(annotation.table), 3) * 100`% of total) transcripts that exceeded the null expectation from the resampling analysis as true positive transcripts for further analyses. Of these transcripts, `r round(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"] / sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), 2) *100`% (`r prettyNum(sigtable[which(sigtable$Coefficient == "Temperature:Colony"), "Number.significant"], big.mark=",")`) had a non-linear relationship with temperature that would likely have been missed with a standard differential expression experiment (e.g. high vs. low temperature). The proportion of responsive transcripts is similar if we focus only on those transcript with a BLAST hit (`r prettyNum(responsive_transcripts_w_BLAST, big.mark = ",")` significant transcripts out of `r prettyNum(nrow(annotation.table.responsive), big.mark = ",")`, `r round(responsive_transcripts_w_BLAST / nrow(annotation.table.responsive), digits = 3)*100`). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] so this estimate is likely a lower bound for the true proportion of transcripts that are thermally responsive.

We used the predicted transcript expression levels to partition transcripts for each species into five expression categories (Fig. 1) which were defined *a priori* to allow us to test predictions derived from three thermal adaptation hypotheses of relative response severity in the two species: **High** transcripts had greatest expression at temperatures > 31°C, **Low** transcripts had greatest expression at temperatures < 10°C, **Intermediate** transcripts had greatest expression between 10 to 30°C, **Bimodal** transcripts had increased expression at both high and low temperatures, while **NotResp** transcripts were those that were not thermally responsive in the focal species but did respond in the other.

@@

In addition, we also examined the specific patterns of shifts from one expression category to another between species. When controlling for overall differences in the number of responsive transcripts, the expression categories of individual transcripts between the two species were not randomly distributed (Stuart-Maxwell test of marginal homogeneity $\chi^2$~4~ = 319, P < 0.001, Fig. S1). Specifically, the two species differed significantly in expression pattern, which captures differences in slope as well as category, for `r prettyNum(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"], big.mark = ",")` (`r round(sigtable[which(sigtable$Coefficient=="Temperature:Colony"), "Number.significant"] / sigtable[which(sigtable$Coefficient=="Total"), "Number.significant"], 2) * 100`%) of the thermally responsive transcripts.

The *enhanced response* and *tolerance* hypotheses make opposing predictions concerning the overlap in response patterns between the two species (Fig. 2). The *enhanced response hypothesis* posits that temperature adaptation uses existing mechanisms for thermal resistance, which should result in significant overlap in response and fewer transcripts shifting expression categories than expected by chance (Fig. 2, left). In contrast, the *tolerance hypothesis* predicts that transcripts involved in active defense will become non-responsive or shift to other expression categories in the better-adapted species (Fig. 2, right). We tested these predictions by examining if the transcripts upregulated in response to the temperature extreme experienced less frequently by each species (cool temperatures for the warm-climate *A. carolinensis*, and warm temperatures for the cool-climate *A. picea*) displayed the same response profile in the species that more frequently experiences those conditions.

Transcripts upregulated at low temperatures in *A. carolinensis* (**Low** and **Bimodal** transcripts) were significantly biased toward this same category and away from other expression categories in *A. picea* (Fig. 3A), suggesting shared response pathways as predicted by the *enhanced response* hypothesis. In contrast, transcripts upregulated in response to high temperatures in *A. picea* (**High** and **Bimodal**) shifted expression categories in *A. carolinensis* (Fig. 3B), primarily to the **Intermediate** category (Fig. 3B). These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.

@@

The gene set enrichment analysis revealed a number of gene groups enriched in each expression category (Table S2). Across both species, there were 9 terms enriched in the **Bimodal** category, including terms involved in stress response (regulation of cellular response to stress, signal transduction by p53 class mediator), cell death (apoptotic signaling pathway) and cellular organization (e.g. protein complex localization). The 6 terms enriched in the **Low** category suggest that proteins undergo structural (e.g protein acylation) and organizational (single-organism organelle organization) changes to tolerate colder temperatures, possibly to maintain membrane fluidity [@ohtsu1998]. The **High** category included only a single enriched GO term, "nicotinamide metabolic process", while the **Intermediate** category had 5 terms including DNA packaging and metabolic process terms.

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *enhanced response* but not the *tolerance hypothesis*, the temperature of induction at low temperatures was significantly higher for the cool-climate *A. picea* than for *A. carolinensis* (`r round(T_on.low.ttest$estimate[1], 1)`°C) than *A. picea* (`r round(T_on.low.ttest$estimate[2], 1)`°C; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 4A), though the temperature of induction did not differ between species for high temperatures (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`).

Similarly, for down-regulated (**Intermediate**) transcripts, we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Consistent with the *tolerance hypothesis*, *A. carolinensis* had greater inertia of gene expression in response to increasing temperatures. The temperature of downregulation for **Intermediate** transcripts was `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`). The difference between species was not significant with decreasing temperatures (*t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 4B).

@@

We tested the *genetic assimilation hypothesis* by comparing the log ratios of relative inducibility to relative baseline expression at the rearing temperature (25°C). If stress-response transcripts have shifted between species from inducible to constitutive expression, there should be a negative relationship between the two. We found no evidence of such a relationship for either temperature extreme: transcripts more upregulated at high temperatures in the cool-climate *A. picea* were not expressed at higher baseline levels in the warm-climate *A. carolinensis* (Fig. 5A). Similarly, transcripts more upregulated at low temperatures in *A. carolinensis* did not show higher baseline levels in *A. picea* (Fig. 5B). In fact, for both comparisons we found a weakly positive relationship between relative inducibility and baseline expression between the two species ($\beta_1$ = 0.31, *P* < 0.001 and ($\beta_1$ = 0.21, *P* < 0.001). In addition, the thermally responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher baseline expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (top-row of Fig. 3B). These transcripts are less likely to be upregulated in any context, consistent with the *tolerance hypothesis*.

@@

Under the *enhanced response* hypothesis, stress-adapted species are hypothesized to induce a stronger and earlier response to extreme conditions. We found evidence for this hypothesis at low temperatures: although the lower thermal limit for *A. picea* is substantially lower than *A. carolinensis*, *A. picea* upregulated responsive transcripts at slightly less extreme temperatures (Fig. 4A). Moreover, the transcripts upregulated in *A. picea* included about half (`r round(tt2["Low", "Low"] / colSums(tt2)["Low"], 2)*100`%) those upregulated in *A. carolinensis* as well as an additional set of `r round(rowSums(tt2)["Low"] - tt2["Low", "Low"], 2)` transcripts (Table 1), enriched for metabolism, organization and translation processes (Table S2). Two non-mutually exclusive hypotheses may explain this pattern. First, surviving prolonged low temperatures, such as would be experienced during overwintering, generally requires advance production of specialized cryoprotectants [@neelakanta2012] and a suite of preparatory physiological modifications [@denlinger2002]. The northern species *A. picea* may induce a greater response to survive the longer winter period. Alternatively, the response to low temperatures may reflect countergradient expression to counteract reduction in enzyme efficiency, and maintain activity as temperature declines [@lockwood2012]. This requirement may be under stronger selection in *A. picea* given the shorter growing season that would necessitate foraging under a broader range of temperatures.

@@

The one hypothesis that did not receive support was the *genetic assimilation hypothesis*, which predicts that exposure to more frequent stressors will select for a shift from inducible to constitutive expression of stress-response transcripts. This constrasts with other recent studies on adaptation in field populations to thermal stress [@barshis2013]. However, in a short-term selection experiment for heat tolerance, Sikkink et al. [-@sikkink2014] also found no evidence for genetic assimilation at the expression level after 10 generations of selection for heat tolerance in *Caenorhabditis remanei*, even though there was a substantial increase in heat tolerance. Both the genetic assimilation and tolerance routes to increasing thermal limits are functionally similar in that they emphasize damage prevention rather than repair. Whether a particular taxon evolves one strategy over another may be related to availability of alternative mechanisms as well as the intensity, frequency and duration of temperature stress in a given environment.

Given the differences in the patterns of thermal responsiveness between species (Fig. 3), it is worth noting a number of similarities. In both species, there were 2 -- 3 times more transcripts upregulated at low than high temperatures (Table 1). The degree of upregulation at low temperatures is surprising given previous studies [@teets2012-1; @vesala2012] that found little transcriptional activity at low temperatures. However, these studies exposed organisms to a few extreme (-10 -- 0°C) temperatures. At these extremes, we also found few upregulated transcripts (Fig. 4A), whereas the peak of low-temperature transcriptional activation occurred near 10°C (Fig. 4). A potential explanation for this pattern is that increased gene expression functions to support elevated metabolism at moderately cold temperatures, as suggested by the metabolic cold adaptation hypothesis [@addo-bediako2002]. The observation that more transcripts were upregulated at low than high temperatures could also be due to stronger selection on upper than lower thermal limits, thereby reducing both genetic variation and gene expression plasticity at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures.

Critical maximum and minimum temperatures are hypothesized to be genetically correlated [@addo-bediako2000; @hoffmann2013], but this was not evident in terms of gene expression in this study. Only ~10% of transcripts upregulated in response to temperature were bimodal, and for both activation and down-regulation, thresholds differed between species at only one temperature extreme (Fig. 4). This suggests that species do not face a fundamental trade-off between these two limits and may be able to shift upper and lower thermal limits independently to match requirements of more seasonally variable environments. 
A major contribution of this study is the construction of a reactionome for gene expression data. Similar approaches have been used in other species [@hodgins-davis2012; @aubin-horth2009], but to our knowledge, none have applied a regression approach to identify a complete list of responsive transcript across an environmental gradient. 
This approach revealed quantitative patterns of temperature response not captured in categorical comparisons. For example, the degree of upregulation at cool (~10°C) but not extreme cold temperatures was missed in previous studies that focused on extreme cold limits, as discussed above. Further, a number of issues have hampered RNA-seq studies to date. Namely, lists of differentially expressed transcripts are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011] and are prone to "storytelling" interpretations [@pavlidis2012]. Our findings are robust to these issues as we focus on the average change in the shape of the reaction norms across many hundreds of responsive transcripts in each species. Although we use gene ontology information to interpret our results, the key findings about differential plasticity of expression between species do not depend on functional annotation.

Moreover, by characterizing responses across thousands of transcripts, the reactionome approach can help to distinguish selection from neutral drift in gene expression [@khaitovich2004; @whitehead2006; @ogasawara2009]. Although we cannot rule out drift as a source of variation for individual transcripts, we would not expect to see systematic differences in expression type categories or critical temperature thresholds as we do here (Fig. 3, Fig. S1). Thus, our method provides an example of how focusing on transcriptome-wide changes in gene expression -- as opposed to identifying lists of differentially-expressed transcripts -- can provide meaningful insight on the process of evolution. It should be noted, however, that although including non-linear relationships between expression and temperature captured a significantly larger range of biologically-relevant responses, it also led to a substantial increase in false positives. Empirical estimation of these rates via randomization tests, combined with robust sampling designs, can help to minimize this bias and focus results on biologically-meaningful gene sets.

@@

We assessed the quality of the transcriptome assembly using the BUSCO program [@simao2015] available from (http://busco.ezlab.org/). BUSCO asseses transcriptome completeness by measuring the number of near-universal single-copy orthologs selected from OrthoDB, using the Arthropod database.

@@

Table S1 provides the annotation, *P*-value, r^2^, adjusted *P*-value, and expression type for the thermally-responsive transcripts in each species.

Table S2 provides the results of the gene set enrichment analysis, showing the enriched gene ontology terms for each species in each thermal response category.

@@

The Trinity transcriptome assembly, FastAnnotator annotation file and Sailfish gene expression quantification files supporting the results of this article are available at the Harvard Forest Archives, HF-113 (http://harvardforest.fas.harvard.edu/data-archive) datasets hf113-38, hf113-41, and hf113-42 (doi:10.6073/pasta/050a6c7c2ff25841b070a0ecab32bc65).

@@

![Fig. 1](../results/expression_types_Fig1.png)

![Fig. 2](../results/predictions_Fig2.png)

![Fig. 3](../results/matched_observations_Fig3.png)

![Fig. 4](../results/crit_temp_regulation_Fig4.png)
